---
title: "practice_1"
format: html
---

## Load libraries

```{r}
library(readr)
library(ggplot2)
```

## Ingest the data from [GitHub SMM635 repo][1]

The dataset comes from Kaggle's [Google Play Store Apps][0] project. Please refer
to the project description for details concerning the data gathering and structuring
aspects.

The project comprises two data tables: app- and review-level data. We will focus 
on the former. 

[0]: https://www.kaggle.com/datasets/lava18/google-play-store-apps

```{r}
data_source <- "https://raw.githubusercontent.com/simoneSantoni/data-viz-smm635/refs/heads/master/data/googleplaystore.csv"
apps <- read_csv(data_source)
```

[1]:https://raw.githubusercontent.com/simoneSantoni/data-viz-smm635/refs/heads/master/data/googleplaystore.csv

## Familiarize with data


### Key Figures

```{r}
spec(apps)
```

### Data Preview

```{r}
apps
```

### Variable Definitions                                                                                                            
#### App                                                                                                                         
- **Type:** Text (String)                                                                                                            
- **Description:** Name of the application                                                                                           
- **Example Values:**                                                                                                                
  - "Photo Editor & Candy Camera & Grid & ScrapBook"                                                                                 
  - "Coloring book moana"                                                                                                            
  - "U Launcher Lite – FREE Live Cool Themes, Hide Apps"                                                                             
  - "Sketch - Draw & Paint"                                                                                                          
- **Missing Values:** Expected to be minimal/none (key identifier)                                                                   
- **Notes:** May contain special characters, emojis, and version information                                                         

#### Category                                                                                                                     
- **Type:** Categorical (String)                                                                                                     
- **Description:** Primary category classification of the app on Google Play Store                                                   
- **Unique Categories:** 33 main categories + some miscoded entries                                                                  
- **Data Quality Issues:** Contains miscoded entries (app names/descriptions appearing as category values)                           
                                                                                                                                     
#### Rating                                                                                                                       
- **Type:** Numeric (Float)                                                                                                          
- **Description:** Average user rating on a scale of 1.0 to 5.0                                                                      
- **Range:** 1.0 - 5.0 (expected)                                                                                                    
- **Example Values:** 4.1, 3.9, 4.7, 4.5                                                                                             
- **Missing Values:** Expected to be present for apps with no ratings                                                                
- **Notes:** Higher values indicate better user satisfaction                                                                         
                                                                                                                                     
#### Reviews                                                                                                                      
- **Type:** Numeric (Integer)                                                                                                        
- **Description:** Total number of user reviews submitted for the app                                                                
- **Example Values:** 159, 967, 87510, 215644                                                                                        
- **Range:** 0 to millions                                                                                                           
- **Notes:** Indicator of app popularity and user engagement                                                                         
                                                                                                                                     
#### Size                                                                                                                         
- **Type:** Text (String with units)                                                                                                 
- **Description:** Size of the application                                                                                           
- **Format:** Number followed by unit (M for megabytes, k for kilobytes)                                                             
- **Example Values:** "19M", "14M", "8.7M", "25M"                                                                                    
- **Special Values:**                                                                                                                
  - "Varies with device" - size depends on device specifications                                                                     
- **Notes:** Requires parsing to convert to numeric format for analysis                                                              
                                                                                                                                     
#### Installs                                                                                                                     
- **Type:** Categorical (String)                                                                                                     
- **Description:** Number of times the app has been installed (in ranges)                                                            
- **Format:** Comma-separated numbers with plus sign                                                                                 
- **Example Values:** "10,000+", "500,000+", "5,000,000+", "50,000,000+"                                                             
- **Notes:** Requires parsing to convert to numeric format; provides order of magnitude rather than exact counts                     
                                                                                                                                     
#### Type                                                                                                                         
- **Type:** Binary Categorical (String)                                                                                              
- **Description:** Whether the app is free or paid                                                                                   
- **Distribution:** Majority of apps are free                                                                                        
- **Notes:** Some data quality issues with miscoded entries                                                                          
                                                                                                                                     
#### Price                                                                                                                        
- **Type:** Text (String with currency symbol)                                                                                       
- **Description:** Price of the app (if paid)                                                                                        
- **Format:** Usually "0" for free apps or "$X.XX" for paid apps                                                                     
- **Example Values:** "0", "$0.99", "$2.99", "$4.99"                                                                                 
- **Notes:**                                                                                                                         
  - Free apps should have "0" value                                                                                                  
  - Requires parsing to remove currency symbol for numeric analysis                                                                  
  - Currency is primarily USD ($)                                                                                                    
                                                                                                                                     
#### Content Rating                                                                                                               
- **Type:** Categorical (String)                                                                                                     
- **Description:** Age group suitability rating for the app                                                                          
- **Standard Values:**                                                                                                               
  - "Everyone" - Suitable for all ages                                                                                               
  - "Everyone 10+" - Suitable for ages 10 and up                                                                                     
  - "Teen" - Suitable for ages 13 and up                                                                                             
  - "Mature 17+" - Suitable for ages 17 and up                                                                                       
  - "Adults only 18+" - Suitable for ages 18 and up                                                                                  
  - "Unrated" - No rating assigned                                                                                                   
- **Example Values:** "Everyone", "Teen"                                                                                             
- **Notes:** Follows Google Play Store's content rating system                                                                       
                                                                                                                                     
#### Genres                                                                                                                      
- **Type:** Text (String, may contain multiple values)                                                                               
- **Description:** Detailed genre classification (may include subcategories)                                                         
- **Format:** Primary genre, sometimes followed by semicolon and secondary genre                                                     
- **Example Values:**                                                                                                                
  - "Art & Design"                                                                                                                   
  - "Art & Design;Pretend Play"                                                                                                      
- **Notes:** More specific than Category field; can contain multiple genres separated by semicolons                                  
                                                                                                                                     
#### Last Updated                                                                                                                
- **Type:** Date (String)                                                                                                            
- **Description:** Date when the app was last updated                                                                                
- **Format:** "Month Day, Year" (e.g., "January 7, 2018")                                                                            
- **Example Values:** "January 7, 2018", "January 15, 2018", "August 1, 2018", "June 8, 2018"                                        
- **Notes:** Can be used to assess app maintenance and developer activity; requires date parsing for analysis                        
                                                                                                                                     
#### Current Ver                                                                                                                 
- **Type:** Text (String)                                                                                                            
- **Description:** Current version of the application                                                                                
- **Format:** Varies widely (semantic versioning, numeric versions, descriptive text)                                                
- **Example Values:** "1.0.0", "2.0.0", "1.2.4", "Varies with device"                                                                
- **Special Values:** "Varies with device" - version depends on device                                                               
- **Notes:** No standardized format across apps                                                                                      
                                                                                                                                     
#### Android Ver                                                                                                                 
- **Type:** Text (String)                                                                                                            
- **Description:** Minimum Android version required to run the app                                                                   
- **Format:** Version number followed by "and up"                                                                                    
- **Example Values:** "4.0.3 and up", "4.2 and up"                                                                                   
- **Special Values:** "Varies with device" - requirement depends on device                                                           
- **Notes:** Indicates app compatibility range      

## Data Visualization

### Univariate Charts

#### Categorical Variables

```{r}
#| fig-cap: Counts of apps by pricing type
#| echo: true

ggplot(data = apps, mapping = aes(x = Type)) +
  geom_bar()
```

```{r}
ggplot(data = apps, mapping = aes(x = Type, fill = Type)) +
  geom_bar() +
  coord_polar(theta = "y") + 
  theme_void()
```

