---
author: Simone Santoni
title: Choosing colors and shapes in ggplot2
date: last-modified
abstract-title: Synopsis
abstract: This notebook illustrates how to change
toc: true
lightbox: true
format:
  html:
    theme: united
    code-fold: true
    code-tools: true
  ipynb: default
  docx: default
  pdf:
    # engine
    pdf-engine: pdflatex
    # page layout
    documentclass: article
    papersize: letter
    geometry:
      - top=2.54cm
      - right=2.54cm
      - bottom=2.54cm
      - left=2.54cm
    # style
    linestretch: 1.1
    indent: true
    # sections
    number-sections: true
    number-depth: 3
    toc: false
    # fonts
    fonts:
      fontenc: true
      fontsize: 11
      #mainfont:TeX Gyre Pagella
      #mainfontoptions:
      #  - Numbers=Lowercase
      #  - Numbers=Proportional
    # colors
    colors:
      linkcolor: blue
      filecolor: red
      citecolor: green
      urlcolor: magenta
      #colorlinks: true
    # captions
    fig-cap-location: bottom
    tbl-cap-location: top
    # footnotes
    links-as-notes: true
    reference-location: document
---

# Notebook setup

## Load libraries

```{r}
library(ggplot2)
library(dplyr)
library(readr)
```

## Load data

The toy dataset we'll use in this notebook is `laptop_price.csv`. It contains information on the price of laptops, as well as the laptops' core featurs. The source for the dataset is https://www.kaggle.com/datasets/muhammetvarl/laptop-price

```{r}
df <- read_csv("~/githubRepos/data-viz-smm635/data/laptops/laptop_price.csv")
df
```

# Colors

## Visual forms' inner color, boarder color, and transparency

In `ggplot2`, it is possible to alter a visual form's default color by passing an optional parameter to the geomtric object at hand. Let's consider a bar chart showing the distribution of laptops across different screen sizes. @fig-base illustrates a chart whose bars exhibit `ggplot2`'s default color. Populating the optional parameter `fill` would alter the chosen visual form's inner color -- see @fig-fill; the optional parameter `colour` affects the visual form's boarder color -- see @fig-fillandboard. It is also possible to regulate the transparency of the chosen color by fixing the optional `alpha` parameter -- see @fig-alpha. Note that the smaller is the scalar value you pass to `alpha`, the more transparent is the visual form -- see @fig-alphaagg.

```{r}
#| label: fig-base
#| fig-cap: "A bar chart with default colors"
p <- ggplot(data = df, mapping = aes(factor(Inches)))
p + geom_bar()
```

```{r}
#| label: fig-fill
#| fig-cap: "A bar chart with custom fill color"
p <- ggplot(data = df, mapping = aes(factor(Inches)))
p + geom_bar(fill = "magenta")
```

```{r}
#| label: fig-fillandboard
#| fig-cap: "A bar chart with custom fill and boarder color"
p <- ggplot(data = df, mapping = aes(factor(Inches)))
p + geom_bar(fill = "magenta", colour = "blue")
```

```{r}
#| label: fig-alpha
#| fig-cap: "A bar chart with adjusted color transparency"
p <- ggplot(data = df, mapping = aes(factor(Inches)))
p + geom_bar(fill = "green", alpha = 0.5)
```
```{r}
#| label: fig-alphaagg
#| fig-cap: "A bar chart with alpha = 0.1"
p <- ggplot(data = df, mapping = aes(factor(Inches)))
p + geom_bar(fill = "green", alpha = 0.1)
```

## Scales

`ggplot2` comes with plenty of [color scales and palettes](https://ggplot2-book.org/scales-colour#brewer-scales) that can help discriminate visually various data groups. Let's suppose to expand on the visualization reported in @fig-boxplot, dealing with the distribution of laptop price across different screen size groups. Specifically, we want to add another dimension to @fig-boxplot to show how laptop prices change across screen and ram size groups. By default, `ggplot2` will use the `hue` color scale see ― @fig-boxplotdefault. To adopt a non-default color scale, the optional argument `scale_color_*` must be populated. In @fig-boxplotbrewer, I adopt a color scale for discrete data, namely [`brewer`](https://ggplot2-book.org/scales-colour#brewer-scales). *Warning*: always ensure to pair discrete (continuous) color scales with discrete (continuous) variables. Otherwise, `ggplot2` will return an error, e.g., ```Discrete values supplied to continuous scale```.

```{r}
#| label: fig-boxplot
#| fig-cap: "Distribution of laptop prices by scree size group"
p <- ggplot(data = df, mapping = aes(x = factor(Inches), y = Price_euros))
p + geom_boxplot()
```
```{r}
#| label: fig-boxplotdefault
#| fig-cap: ""
p <- ggplot(data = df, mapping = aes(x = factor(Inches), y = Price_euros))
p + geom_boxplot(aes(colour = Ram))
```

```{r}
#| label: fig-boxplotbrewer
#| fig-cap: ""
p <- ggplot(data = df, mapping = aes(x = factor(Inches), y = Price_euros))
p + geom_boxplot(aes(colour = Ram)) + scale_color_brewer(palette = "Paired")
```

# Shapes

In data visualization, shapes can play a role similar to colors, by representing further data dimensions. For example, @fig-colors and @fig-shapes use color and shapes, respectively, to denote two different data series regarding Apple and Lenovo laptops.

```{r}
#| label: fig-colors
#| fig-cap: "Colors denote companie"
cols <- c("Company", "Inches")
apple_lenovo <- df |> filter(Company == "Apple" | Company == "Lenovo")
ave <- apple_lenovo |> group_by(across(all_of(cols))) |> summarize(ave_price = mean(Price_euros))
p <- ggplot(data = ave, mapping = aes(x = factor(Inches), y = ave_price, color = factor(Company)))
p + geom_point()
```
```{r}
#| label: fig-shapes
#| fig-cap: "Shapes denote companies"
p <- ggplot(data = ave, mapping = aes(x = factor(Inches), y = ave_price, shape = factor(Company)))
p + geom_point()
```

At the same time, one may want to adopt a non-default shape across all data series. That would be the case of @fig-changingshapes, in which shape '5' ― an empty circle ― replaces `ggplot2`'s default shape. 

```{r}
#| label: fig-changingshapes
#| fig-cap: "A `geom_point()` with a non-default shape"
p <- ggplot(data = ave, mapping = aes(x = factor(Inches), y = ave_price, color = factor(Company)))
p + geom_point(shape = 5)
```
