---
author: Simone Santoni
title: Choosing colors and shapes in ggplot2
date: last-modified
abstract-title: Synopsis
abstract: This notebook illustrates how to change 
format:
  html:
    fig-width: 6
    fig-height: 4
    code-fold: true
---

## Notebook setup

### Load libraries

```{r}
library(ggplot2)
library(readr)
```

### Load data

The toy dataset we'll use in this notebook is `laptop_price.csv`. It contains information on the price of laptops, as well as the laptops' core featurs. The source is https://www.kaggle.com/datasets/muhammetvarl/laptop-price

```{r}
df <- read_csv("/home/simone/githubRepos/data-viz-smm635/data/laptops/laptop_price.csv")
df
```

## Colors

In `ggplot2`, it is possible to alter a visual form's default color by passing an optional parameter to the geomtric object at hand. Let's consider a bar chart showing the distribution of laptops across different screen sizes. @fig-base illustrates a chart whose bars exhibit `ggplot2`'s default color.

```{r}
#| label: fig-base
#| fig-cap: "A bar chart with default colors"
p <- ggplot(data = df, mapping = aes(factor(Inches)))
p + geom_bar()
```

## Shapes

