---
author: Simone Santoni
date: last-modified
title: Changing `ggplot` theme
abstract: This notebooks shows how to change a `ggplot2` plot's theme. It considers both the built-in themes in `ggplot2` and some 'extra' themes available in `ggplot2` extensions.
format: 
  html:
    fig-width: 6
    fig-height: 4
    code-fold: true
---

## Notebook setup

### Load packages

```{r}
library(ggplot2)
library(readr)
library(hrbrthemes)
library(ggthemes)
```
### Ingest the data

```{r}
getwd()
df <- read_csv("../../data/hrDataset/HR_capstone_dataset.csv")
```

### Data preview

```{r}
summary(df)
```

## Applying multiple themes to a sample plot: `number_project` vs `average_montly_hours`

### Default theme

```{r}
p <- ggplot(data = df, mapping = aes(x = factor(number_project), y = average_montly_hours, colour = salary))
p + geom_boxplot()
```

### `ggplot2` alternatives

`ggplot2` provides the following set of alternative themes:

+ theme_grey() 
+ theme_gray() 
+ theme_bw() 
+ theme_linedraw() 
+ theme_light() 
+ theme_dark() 
+ theme_minimal() 
+ theme_classic() 
+ theme_void() 
+ theme_test()

`ggplot2` themes can be applied *ad hoc*, like in @fig-thememin, or using the `theme_set()` function, which changes the default theme.

```{r}
#| label: fig-thememin
#| fig-cap: "Distribution of average monthly hours by number of projects and salary level"
p + geom_boxplot() + theme_minimal()
```

*Experiment* with `ggplot2` themes to discover the theme that you want to pass to `theme_test()`.

### `ggthemes` alternatives

The `ggplot2` community has also created some popoular extensions providing 'extra' themes. [`ggthemes`](https://github.com/jrnold/ggthemes) is one of these extensions. This time we change the default theme to `theme_economist()`. As @fig-theecon shows, the colors of the visual forms have not changed though (they are `ggplot2`'s default). To adopt a theme and its colors, we have to pass a further option, as per @fig-fulltheecon

```{r}
#| label: fig-theecon
#| fig-cap: "An illustration of `ggthemes`' `theme_economist()`"
theme_set(theme_economist())
p + geom_boxplot()
```

```{r}
#| label: fig-fulltheecon 
#| fig-cap: "An illustration of `ggthemes`' `theme_economist()`"
p + geom_boxplot() + scale_color_economist()
```

### `hrbrthemes` alternatives

`hrbrthemes` is yet another extension offering extra themes for `ggplot2`. A distinctive feature of this extension is the selection of fonts associatied with the various themes. For example, the theme `theme_ipsum` comes with seven alternative fonts.


```{r}
#| label: fig-thmeipsum
#| fig-cap: "An illustration of `hrbrthemes`'s `theme_ipsum()`"
theme_set(theme_ipsum())
p + geom_boxplot()
```

```{r}
#| label: fig-thmeipsumrc
#| fig-cap: "An illustration of `hrbrthemes`'s `theme_ipsum_rc()`"
theme_set(theme_ipsum_rc())
p + geom_boxplot()
```

`hrbrthemes` offers 'dark' themes too (see @fig-hrdark)

```{r}
#| label: fig-hrdark
#| fig-cap: "An illustration of `brbrthemes`'s `theme_ft_rc`"
theme_set(theme_ft_rc())
p + geom_boxplot() + scale_color_ft()
```
